---
title: "nmIRT"
author: "Rikard Nordgren"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{nmIRT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

nmIRT as an R package for generation and diagnostics of IRT NONMEM models. This document displays all different features of the package. For a tutorial introduction please read the introduction vignette.

## Scales

A scale object is a representation in R of a certain scale. It is a collection of items that each has an item number, a type (currently ordered categorical and binmary are supported) and a set of levels. A scale object can be created in multiple ways.

### Using a built in scale

```{r, fig.show='hold'}
library(nmIRT)
scale <- load_predefined_scale("MDS-UPDRS")
```
To find out which scales are available
```{r, fig.show='hold'}
list_predefined_scales()
```

### Getting an overview of a scale
A summary table of all items of a scale can be generated
```{r}
summary(scale)
```

There is also the option of getting just the item numbers available or all that are of a certain type
```{r, fig.show='hold'}
all_items(scale)
items_by_type(scale, item_type$binary)
```


### Generating a scale from a data.frame
```{r, fig.show='hold'}
df <- data.frame(ID=c(1,1,1,2,2,2), ITEM=c(1,2,3,1,2,3), DV=c(0,2,4,1,2,1))

scale <- create_scale_from_df(df)
```

By default the item column is called ITEM and the dv column is called DV, but this can be changed with options
```{r, eval=FALSE, fig.show='hold'}
scale <- create_scale_from_df(df, item='ITEM', dv='DV')
```

Note that if the dataset has an MDV column it will be used to filter out missing DVs.

### Generating a scale from a CSV file
```{r, eval=FALSE, fig.show='hold'}
scale <- create_scale_from_csv("mydata.csv")
```

### Load scale from a scale definition file
More information on how to work with scale definition files can be found in [a separate vignette](scale_defintion_files.html)

```{r, eval=FALSE, fig.show='hold'}
scale <- load_scale("myscale.yaml")
```

### Save a scale to a scale definition file
```{r, eval=FALSE, fig.show='hold'}
scale <- save_scale(scale, "myscale.yaml")
```


### Check a dataset against a scale
```{r, eval=FALSE}
check_data(dataset, scale)
```
This function will check if there are items or levels missing from the dataset that are present in the scale or vice versa.


## Modifying a scale

### Create subscales using categories
Items of a scale can belong to zero or more categories. These categories simplifies the creation of subscales. The following example will create a subscale of the items belonging to the "motor" category.

```{r, fig.show='hold', eval=FALSE}
subscale <- create_subscale(scale, c("motor"))
```

It is also possible to simply list all items in a certain combination of categories.

```{r}
items_in_categories(load_predefined_scale("MDS-UPDRS"), c("motor", "tremor"))
```

### Remove items from a scale
```{r, fig.show='hold', eval=FALSE}
new_scale <- remove_items(scale, 1:14)
```

## Creating a model object from a scale
```{r, fig.show='hold'}
model <- irt_model(scale)
```

### Attach a dataset to a model
```{r, eval=FALSE}
model <- set_dataset(model, "my.dta")
```

It is also possible to not use the full path of the dataset in $DATA
```{r, eval=FALSE}
model <- set_dataset(model, "my.dta", use_path=FALSE)
```
This is useful when creating and running the model on different machines.


### Generate the NONMEM code from a model object
```{r, fig.show='hold', eval=FALSE}
scale <- create_scale_from_dataset(df)
model <- irt_model(scale)
print_model_code(model)
```

### Write the NONMEM code to a file
```{r, eval=FALSE}
save_model_code(model, "run1.mod")
```

## Make changes to a model

### Add a simulation task
```{r, eval=FALSE}
model <- add_simulation(model, nsim=50)
```

This function also has the optional argument ```options``` that takes a string that will be added verbatim to the $SIMULATION.

### Consolidate levels by fixing parameters
It is possible to consolidate levels in the upper end of the level range by fixing the corresponding DIF parameter to a high value and keeping the level in the model. The benefit of this is to facilitate using different datasets for the same model by only changing the initial estimates. This is done on the model object:
```{r, eval=FALSE}
model <- consolidate_levels(model, 23, c(4))
```

This will set the probability of level 4 to 0 by fixing DIF4 to a high value. The item number can be a list of items to consolidate the same levels for.

A convenience function to get the count for each item and level is available:
```{r, eval=FALSE}
item_level_count(model_or_dataframe)
```

To consolidate all levels that does not occur frequently in the dataset use
```{r, eval=FALSE}
count <- 5
model <- consolidate_levels_below(model, count)
```
The count is the minimum number of occurences for any level in the dataset to have it consolidated. Note that the model has to have an added dataset before using this function.


### Set the run number
```{r, eval=FALSE}
model <- set_run_number(model, 28)
```

This will set the run number to 28, which will add this number to the names of all table files generated by the model. If a number was not set the default of 1 would be used.

### Add options to $ESTIMATION
```{r, eval=FALSE}
model <- add_estimation_options(model, "SIGDIGITS=4 SADDLE_RESET=1")
```

## Working with parameters

### Getting parameter names

```{r}
item_parameter_names(model, 1)
```

### Initial estimates
There are multiple ways of setting the initial estimates for the item parameters in an irt model. This is the order of priority of how the initial estimates are determined:

1. User defined value from ```set_initial_estimates```
2. 50 if the parameter was consolidated
3. Taken from the published model in the scale
4. A fall back value of:
       Binary: DIS=1, DIF=0.1, GUE=0.01
       Ordered categorical: DIS=1, DIFn=$-3 + \frac{6n}{n_{cats}}$ (not lower than 0.1)

### Providing user defined initial estimates
The most straight forward way of setting some initial estimates is to use the function ```set_initial_estimates```.

For example
```{r, eval=FALSE}
model <- set_initial_estimates(model, 1:13, c("DIF2", "DIF3", "DIF4"), 0.1)
```
will set all the initial estimates for the DIF2, DIF3 and DIF4 parameters for items 1 through 13 to 0.1

A vector of initial estimates can also be supplied and these will then be broadcast to each parameter. For example
```{r, eval=FALSE}
model <- set_initial_estimates(model, 1:3, c("DIF2", "DIF3"), c(0.1, 0.2))
```
will set the initial estimates for DIF2 and DIF3 for items 1, 2 and 3 to 0.1 for the DIF2:s and 0.2 for the DIF3:s

### Fixing parameters

It is possible to fix certain item parameters of a model.

For example
```{r, eval=FALSE}
model <- fix_item_parameters(model, items_in_categories(scale, "motor"), item_parameter_names(scale, 1))
```
will fix all parameters in the motor category

### Ignoring items

An item can be ignored in a model. This is done by keeping the THETAs for the item parameters as placeholders to be able to keep the same THETA numbers. The items will also be ignored in $DATA.

```{r, eval=FALSE}
model <- ignore_items(model, 1:13)
```

### Getting an overview of initial estimates

```{r}
inits <- list_initial_estimates(model)
head(inits, n=12)
```

### Estimate initial estimates given the data
The mirt package can be used to estimate the initial estimates given the dataset.

```{r, eval=FALSE}
item_prm <- estimate_item_parameters(model)
model <- set_initial_estimates_table(model, item_prm)
```

## Latent variable models

An IRT model can describe one or more latent variables. Each latent variable can be modelled differently. A latent variable model can be connected to one or more items. Either a pre-defined model can be used:

```{r}
model <- add_lv_model(model, lv_model_type="constant", items=1:14)
```

or a custom model

```{r}
model <- reset_lv_models(model)
lv_model_code <- c("PSI = THETA(1) + ETA(1)")
model <- add_custom_lv_model(model, lv_model_code, items=items_in_categories(model, "motor"))
```

The latent variable models have to be reset before changing them as illustrated above. To simplify setting the items ```all_items```, ```items_by_type``` and ```items_in_categories``` can be used.


## Diagnostic plots

The diagnostic plots can be created after an IRT model has been run. They will take as input data frames containing output from NONMEM.

### Item characteristic curves

To generate a set of item characteristic curves

```{r, eval=FALSE}
df <- read.table("item_parameters_tab1", skip=1, header=T, sep=",")
plots <- icc_plots(df, model, items_per_page=6)
print(plots)
```

### Mirror plot

The mirror plot gives an overview of the frequencies of levels in the dataset and compares it to frequencies in simulations.

```{r, eval=FALSE}
origdata <- read.table("mydata.dta")
simdata <- read.table("simulation_tab1", skip=1, header=T, sep=",")
plots <- mirror_plots(origdata, model, simdata, nrow=4, ncol=5)
print(plots)
```

### Correlation plot

The correlation plots give a heatmap of correlations between residuals.

```{r, eval=FALSE}
psi.estimates <- read.table("psi_tab1", skip=1, header=T)
plot <- correlation_plot(psi.estimates)
print(plot)
```