---
title: "nmIRT"
author: "Rikard Nordgren"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{nmIRT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

nmIRT as an R package for generation and diagnostics of IRT NONMEM models.

## Scales

A scale object is a representation in R of a certain scale. It is a collection of items that each has an item number, a type (currently only ordered categorical is supported) and a set of levels. A scale object can be created in multiple ways. 

### Using a built in scale

```{r, fig.show='hold'}
library(nmIRT)
scale <- predefined_scale("MDS-UPDRS")
```

### Generating a scale from a data.frame
```{r, fig.show='hold'}
df <- data.frame(ID=c(1,1,1,2,2,2), ITEM=c(1,2,3,1,2,3), DV=c(0,2,4,1,2,1))
scale <- scale_from_dataset(df)
```

By default the item column is called ITEM and the dv column is called DV, but this can be changed with options
```{r, eval=FALSE, fig.show='hold'}
scale <- scale_from_dataset(df, item='ITEM', dv='DV')
```

Note that if the dataset has an MDV column it will be used to filter out missing DVs.

A summary of the scale can be displayed with
```{r, fig.show='hold'}
str(scale)
```
Note that what is displayed here is the internal structure of the scale object and that it can be subject to change.


### Generating a scale from a dataset file
```{r, eval=FALSE, fig.show='hold'}
scale <- scale_from_dataset("mydata.csv")
```

## Creating a model object from a scale
```{r, fig.show='hold'}
model <- irt_model(scale)
```

## Generate the NONMEM code from a model object
```{r, fig.show='hold'}
cat(render(model))
```
